---
title: "Part 2.2"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r}
library("ctsmTMB")
library("lubridate")
library("tidyverse")
library("dplyr")
library("ggplot2")
library("patchwork")
```

First step, we explore the data.

```{r}
data <- read.csv('/Users/sunjiawen/Desktop/Advanced ts/02427_CE03_Rainfall_runoff_exercise_data/ex2_overflow.csv')

head(data)
```

Then we need to convert data format as the question demanding.

```{r}
df <- data.frame(
  t = as.numeric(difftime(data$timestamp, min(data$timestamp), units = "hours")),
  U = as.numeric(data$rainfall),
  y = as.numeric(data$stormwater)
)

```

Try to plot two graphs as Part2 shows

```{r}
par(mfrow = c(2, 1), mar=c(4,4,2,1))

# Runoff Plot
plot(df$t, df$y, type="l", lwd=2,
     xlab="", ylab="Runoff", main="")

# Rainfall Plot
plot(df$t, df$U, type="l", lwd=1,
     xlab="Time", ylab="Rainfall", main="")

```

They are the same as the demand.

Now, we define the parameters：

# 2.2.1 Linear Reservoir Model (n = 2)

In this section, we derive the continuous-time stochastic differential
equation (SDE) formulation of a **linear reservoir model** with **three
states**, corresponding to three connected reservoirs. The model
represents how rainfall is transformed into stormwater runoff.

## Model Structure

We consider a chain of three connected linear reservoirs:

$$
U(t) \rightarrow X_1(t) \rightarrow X_2(t) \rightarrow X_3(t) \rightarrow y(t)
$$

Where:

-   $U(t)$: rainfall input\
-   $X_1(t), X_2(t), X_3(t)$: storage levels of the three reservoirs\
-   $y(t)$: observed stormwater runoff\
-   Each reservoir releases water downstream proportionally to its
    current storage level.

This assumption yields a **linear system**, which can be described by a set of SDEs.

------------------------------------------------------------------------

## Relationship Between n and the Number of Reservoirs

In the notation used in the course notes, the parameter $n$ is defined
as:

$$
n = \text{(number of reservoirs)} - 1
$$

Since we use **3 reservoirs**, we obtain:

$$
n = 3 - 1 = 2
$$

Therefore, every inter-reservoir flow coefficient takes the form:

$$
\frac{n}{K} = \frac{2}{K}
$$

where $K$ is a time-scale parameter controlling how quickly water moves
through the system.

------------------------------------------------------------------------

## Stochastic Differential Equations

Below we formulate the SDEs that describe the reservoir dynamics.

### **State 1 (receives rainfall input)**

$$
dX_1 =
A\,U_t\,dt 
- \frac{2}{K} X_1\,dt
+ \sigma\, dW_1(t)
$$

State 1 receives rainfall scaled by the catchment area parameter $A$,
and releases water to State 2 at rate $\frac{2}{K}$.

------------------------------------------------------------------------

### **State 2 (receives water from** $X_1$)

$$
dX_2 =
\frac{2}{K} X_1\,dt
- \frac{2}{K} X_2\,dt
+ \sigma\, dW_2(t)
$$

------------------------------------------------------------------------

### **State 3 (receives water from** $X_2$)

$$
dX_3 =
\frac{2}{K} X_2\,dt
+ \sigma\, dW_3(t)
$$

This is the final reservoir before the observed outflow.

------------------------------------------------------------------------

## Measurement Equation

The measured stormwater runoff $y_t$ is assumed to equal the outflow of
the last reservoir:

$$
y_t = X_3(t) + e_t
$$

where $e_t$ represents measurement noise.

------------------------------------------------------------------------

Next, we will translate these equations into R code and estimate the
parameters from data.

```{r}

# Linear reservoir model with 3 states (n = 2)
m <- ctsmTMB$new()


# STATE 1
m$addSystem(
  dX1 ~ A * U * dt - (2/K) * X1 * dt + sigma * dw
)

# STATE 2
m$addSystem(
  dX2 ~ (2/K) * X1 * dt - (2/K) * X2 * dt + sigma * dw
)

# STATE 3
m$addSystem(
  dX3 ~ (2/K) * X2 * dt + sigma * dw
)

# y observes X3
m$addObs(
  y ~ X3
)

# Add measurement noise variance (sig_e)
m$setVariance(
  y ~ sigma_y^2
)

# Add Input(Rainfall)
m$addInput(U)

# Set Initial States
m$setInitialState(list(c(df$U[1], df$U[1], df$y[1]), 1e-1 * diag(3)))


# Parameter initialization
m$setParameter(
  A = c(initial = 1, lower = 0, upper = 500),
K = c(initial = 20, lower = 1e-3, upper = 500),
sigma = c(initial = 0.1, lower = 1e-10, upper = 30),
sigma_y = c(initial = 0.1, lower = 1e-10, upper = 30)
)

# Predict the model
fit_lin <- m$estimate(df, method = "ekf", compile = TRUE)
pred_lin <- m$predict(df, k.ahead = 10)

```

```{r}
summary(fit_lin)
```


## Visualization

```{r}
s10 <- pred_lin$states %>% filter(k.ahead == 10)

ggplot() +
  geom_line(aes(x = s10$t.j, y = s10$X3, color = "10-step Prediction"),
            size = 1.1) +

  geom_line(aes(x = df$t, y = df$y, color = "Observed"),
            size = 0.9) +

  scale_color_manual(values = c(
    "Observed" = "black",
    "10-step Prediction" = "red"
  )) +

  labs(title = "10-step Ahead Prediction vs Observed",
       x = "Time", y = "Stormwater",
       color = "") +
  
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")



```





```{r}
# Residuals extraction
residuals <- fit_lin$residuals$residuals$y

residuals

ggplot(data.frame(t = df$t, residuals = residuals), aes(x = t, y = residuals)) +
  geom_line() +
  labs(title = "Residuals over Time", x = "Time", y = "Residuals", color = "blue") +
  theme_minimal()

# QQ plot
qqnorm(residuals)
qqline(residuals, col = "red")

# ACF (LACF)
acf(residuals)

# PACF (PLACF)
pacf(residuals)
```




### Summary:


# 2.2.2 Introduce overflow into the system

```{r}

# Create a model for 2.2.2
m2 <- ctsmTMB$new()

#----------------------------------
# STATE EQUATIONS
#----------------------------------

# STATE 1
m2$addSystem(
  dX1 ~ A * U * dt 
       - (2/K) * X1 * dt 
       + sigma * dw1
)

# STATE 2
m2$addSystem(
  dX2 ~ (2/K) * X1 * dt 
       - (2/K) * X2 * dt 
       + sigma * dw2
)

# STATE 3 with overflow
m2$addSystem(
  dX3 ~ (2/K) * (1 / (1 + exp(-alpha * (X3 - beta)))) * X2 * dt
       + sigma * dw3
)

#----------------------------------
# OBSERVATION EQUATIONS
#----------------------------------

m2$addObs(
  y ~ X3
)

# Measurement noise
m2$setVariance(
  y  ~ sigma_storm^2
)

#----------------------------------
# INPUT
#----------------------------------
m2$addInput(U)

#----------------------------------
# INITIAL STATE
#----------------------------------
m2$setInitialState(list(c(df$U[1], df$U[1], df$y[1]), 1e-1 * diag(3)))


#----------------------------------
# PARAMETERS
#----------------------------------
m2$setParameter(
  A = c(initial = 1, lower = 0, upper = 500),
  K = c(initial = 1, lower = 1e-3, upper = 500),

  alpha = c(initial = 1, lower = 0.01, upper = 1000),
  beta  = c(initial = 0.5, lower = 0.01, upper = 1000),

  sigma = c(initial = 0.1, lower = 1e-10, upper = 30),
  sigma_storm  = c(initial = 0.1, lower = 1e-10, upper = 30)
)

#----------------------------------
# FIT THE MODEL
#----------------------------------
fit <- m2$estimate(df, method = "ekf", compile=T)

# Predictions
pred <- m2$predict(df, k.ahead = 10)

```

```{r}
summary(fit)
```

```{r}
# Extract the 1-step ahead (prior) estimates\
s <- pred$states %>% filter(k.ahead == 10)

ggplot() +
  geom_ribbon(aes(x = s$t.j,
                  ymin = s$X3 - 2*sqrt(s$var.X3),
                  ymax = s$X3 + 2*sqrt(s$var.X3)),
              fill = "grey80", alpha = 0.6) +

  geom_line(aes(x = s$t.j, y = s$X3, color = "10-step Prediction"),
            size = 1.1) +

  geom_line(aes(x = df$t, y = df$y, color = "Observed"),
            size = 0.9) +

  scale_color_manual(values = c(
    "Observed" = "black",
    "10-step Prediction" = "red"
  )) +
  
  labs(title = "10-step Ahead Prediction vs Observed",
       x = "Time",
       y = "Stormwater",
       color = "") +
  
  theme_minimal() +
  theme(legend.position = "top")

```


```{r}
residuals_2 <- fit$residuals$residuals$y

ggplot(data.frame(t = df$t, residuals = residuals_2), 
       aes(x = t, y = residuals_2)) +
  geom_line() +
  labs(title = "Residuals",
       x = "Time", y = "Residual") +
  theme_minimal()

# QQ plot
qqnorm(residuals_2)
qqline(residuals_2, col = "red")

# ACF (LACF)
acf(residuals_2)

# PACF (PLACF)
pacf(residuals_2)

```


report the log-likelihood and all parameters

```{r}
logLik_lin <- -fit_lin$nll
logLik_lin

logLik_sig <- -fit$nll
logLik_sig

```




## 2.2.3
Report the correlation matrix for the parameters.
```{r}
cor_matlin <- cov2cor(fit_lin$cov.fixed)
cor_matlin
cor_matlin2 <- cov2cor(fit$cov.fixed)
cor_matlin2

```
```{r}
fit_lin$par.fixed
fit$par.fixed
```



